Coherent optical creation of a single molecule -- Supplemental material

### 2020-12-15 sm.tex
Update on Overleaf.

\newcommand{\harvardphysics}{\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts 02138, USA}}
\newcommand{\harvardccb}{\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts 02138, USA}}
\newcommand{\cua}{\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts 02138, USA}}
\newcommand{\gradstudent}{
  \harvardphysics
  \harvardccb
  \cua
}

@@

\begin{document}
\title{Coherent optical creation of a single molecule -- Supplemental material}
\author{Yichao~Yu}
\email{yichaoyu@g.harvard.edu}
\gradstudent
\author{Kenneth~Wang}
\gradstudent
\author{Jonathan~D.~Hood}
\affiliation{Department of Chemistry, Purdue University, West Lafayette, Indiana, 47906, USA}
\author{Lewis~R.~B.~Picard}
\gradstudent
\author{Jessie~T.~Zhang}
\gradstudent
\author{William~B.~Cairncross}
\harvardccb
\harvardphysics
\cua
\author{Jeremy~M.~Hutson}
\affiliation{Joint Quantum Centre Durham-Newcastle, Department of Chemistry, Durham University, Durham, DH1 3LE, United Kingdom}
\author{Till Rosenband}
\harvardphysics
\author{Kang-Kuen~Ni}
\email{ni@chemistry.harvard.edu}
\harvardccb
\harvardphysics
\cua

@@

In our system, the separation in energy between the initial and target state is much smaller than the single photon detuning $\Delta$ from the intermediate state.
Thus, there is significant cross coupling for the scattering rate and light shift, where each state is coupled to the excited state by the power in both beams.
Under these conditions, the scattering rate and light shift is proportional to the total power $P_{\text{tot}} = P_{\rm m} + P_{\rm a}$, where $P_{\rm m} (P_{\rm a})$ is the power in the beam that addresses the molecular (atomic) state.
The Raman Rabi frequency is proportional to $\sqrt{P_{\rm m}P_{\rm a}} $.
With a fixed total power and thus fixed scattering rate, the Raman Rabi frequency is maximized when $ P_{\rm m} = P_{\rm a} = P_{\text{tot}}/2$.

@@

With multiple intermediate states, the total Raman Rabi frequency, scattering rate, and light shift become a sum over all these possible states. In our model, we consider all vibrational states of $ \mathrm{c^3\Sigma^+(\Omega = 1)} $ and states in the atomic continuum as intermediate states. The Rabi frequency between two states is proportional to the electronic dipole moment, arising from the electronic wavefunctions, and the overlap of the vibrational wavefunctions. Under the assumption that the electronic dipole moment $ d $ is the same for all transitions between the ground state and the excited electronic state, the total Raman Rabi frequency becomes:

@@

  \Omega_{\rm R} = \frac{d^2E^2}{2\hbar^2} \displaystyle\sum_{v'} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}}

@@

where $\ket{a}$ ($\ket{m}$) are the relative coordinate wavefunctions for the trapped atomic (target molecular) state, and $\ket{v'}$ is the relative-coordinate wavefunction for the excited state which includes both bound molecular states and the atomic continuum. $\Delta_{v'}$ is a single photon detuning with respect to the $\ket{v'}$ state, $ E $ is the electric field for the power in the single beam given a waist size. We can divide this sum into a sum over all the bound molecular states, and a sum over atomic continuum states.

@@

  \Omega_{\rm R} = \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}} +  \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' > v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}}

@@

where $ v'_{\text{max}}$ corresponds to the excited molecular state, which is most weakly bound. The first term can be calculated numerically given an excited state potential. To calculate the second term, we first make an assumption that we are far enough detuned from the contributing continuum states that they are all at approximately the same detuning, $ \Delta_{\text{thresh}}$. This allows us to pull it out of the sum, and we obtain:

@@

  \Omega_{\rm R} \approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}} +  \frac{d^2E^2}{2\hbar^2\Delta_{\text{thresh}}}\displaystyle\sum_{v' > v'_{\text{max}}} \braket{a|v'}\braket{v'|m}  \label{rabifreq}

@@

From requiring orthogonality,

@@

  \displaystyle\sum_{v'} \braket{a|v'}\braket{v'|m} = \braket{a|m} = 0

@@

  \displaystyle\sum_{v'> v'_{\text{max}}} \braket{a|v'}\braket{v'|m} = -\displaystyle\sum_{v' \leq v'_{\text{max}}} \braket{a|v'}\braket{v'|m}

@@

  \Omega_{\rm R} \approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}} -  \frac{d^2E^2}{2\hbar^2\Delta_{\text{thresh}}}\displaystyle\sum_{v' \leq v'_{\text{max}}} \braket{a|v'}\braket{v'|m}

@@

We can account for the atomic continuum in the scattering rate in a similar way.

@@

  \Gamma_{\rm s} &= \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}}\Gamma_{v'} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}^2} + \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' > v'_{\text{max}}}\Gamma_{v'} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}^2} \\
           &\approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}}\Gamma_{v'} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}^2} + \frac{d^2E^2\Gamma}{2\hbar^2\Delta_{\text{thresh}}^2}\displaystyle\sum_{v' > v'_{\text{max}}}\braket{m|v'}\braket{v'|m}

@@

  \displaystyle\sum_{v' > v'_{\text{max}}}\braket{m|v'}\braket{v'|m} = 1 - \displaystyle\sum_{v' \leq v'_{\text{max}}}\braket{m|v'}\braket{v'|m}

@@

  \Gamma_{\rm s} \approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}}\Gamma_{v'} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}^2} +  \frac{d^2E^2\Gamma}{2\hbar^2\Delta_{\text{thresh}}^2}\left(1 - \displaystyle\sum_{v' \leq v'_{\text{max}}}\braket{m|v'}\braket{v'|m}\right)

@@

A very similar result holds for the light shift and is written here for completeness

\begin{equation}
  \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}} +  \frac{d^2E^2}{2\hbar^2\Delta_{\text{thresh}}}\left(1 - \displaystyle\sum_{v' \leq v'_{\text{max}}}\braket{m|v'}\braket{v'|m}\right)
\end{equation}

\section{Computational methods for bound states and scattering}

@@

\section{Calculating Matrix Elements with Coupled Channel Ground State}
To calculate Rabi frequencies for transitions to the eight excited molecular potentials, we
obtain their wavefunctions from a single-channel calculation \cite{Fattal1996} using the potential energy curves described in section \ref{sm:excited_states}. We
then calculate overlap integrals between these functions and either the triplet ($S=1$) or singlet ($S=0$) components of the
coupled-channel wavefunctions for either the target molecular state or the confined atomic state.
These calculations assume that the intermediate states have pure singlet or triplet character and neglect its
nuclear spin character, which is unknown. These calculations include theoretical electronic transition dipole moments~\footnote{Private communication with R. Gonzalez-Ferez and R. Moszynski}.

@@

In this calculation, there is a matrix element, $\Omega_{\mathrm{a}i}, \Omega_{\mathrm{m}i}$ between each channel $ i $ of the coupled channel ground state and an excited state. The total Raman Rabi frequency $ \Omega_{\rm R}$, light shift $ \delta$ and off-resonant scattering $ \Gamma_{\rm s} $ contribution from this single excited state is calculated as

\begin{align}
  \Omega_{\rm R} &= \displaystyle\sum_{i} \frac{\Omega_{\mathrm{a}i} \Omega_{\mathrm{m}i}}{2\Delta} \\
  \delta &= \displaystyle\sum_{i} \frac{\Omega_{\mathrm{a}i}^2 - \Omega_{\mathrm{m}i}^2}{2\Delta} \\
  \Gamma_{\rm s} &= \displaystyle\sum_{i} \Gamma_{\rm e} \frac{\Omega_{\mathrm{a}i}^2 + \Omega_{\mathrm{m}i}^2}{2\Delta^2}
\end{align}

where $ \Delta $ is the single photon detuning and $ \Gamma_{\rm e} $ is the excited state linewidth. The experiment, through measuring the detuning dependence of the Raman Rabi frequency and the resonance frequency, extracts a single $ \Omega_{\rm m} $ and $ \Omega_{\rm a} $. In order to compare to the theory, we introduce effective $ \Omega_{\rm a}', \Omega_{\rm m}', \Gamma_{\rm e}'$, so that $ \Omega_{\rm R} = \Omega_{\rm a}'\Omega_{\rm m}'/(2\Delta) $, $\delta = (\Omega_{\rm a}'^2 - \Omega_{\rm m}'^2)/(2\Delta) $, and $\Gamma_{\rm s} = \Gamma_{\rm e}'(\Omega_{\rm a}'^2 + \Omega_{\rm m}'^2)/(2\Delta^2) $. The effective Rabi frequencies for the $v' = 0$ state of $ c^3\Sigma^+_1$ and their ratio are reported in Table~\ref{tab:sm} at $3.75~\mathrm{mW}$ of tweezer power.

>>>>

\newcommand{\harvardphysics}{\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts 02138, USA}}
\newcommand{\harvardccb}{\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts 02138, USA}}
\newcommand{\cua}{\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts 02138, USA}}
\newcommand{\gradstudent}{
  \harvardphysics
  \harvardccb
  \cua
}
\newcommand{\itamp}{\affiliation{ITAMP, Harvard-Smithsonian Center for Astrophysics, Cambridge, Massachusetts 02138, USA}}

@@

\begin{document}
\title{Coherent optical creation of a single molecule -- Supplemental material}
\author{Yichao~Yu}
\thanks{Y.Y.~and K.W.~contributed equally to this work.}
\email{yichaoyu@g.harvard.edu}
\gradstudent
\author{Kenneth~Wang}
\thanks{Y.Y.~and K.W.~contributed equally to this work.}
\gradstudent
\author{Jonathan~D.~Hood}
\affiliation{Department of Chemistry, Purdue University, West Lafayette, Indiana, 47906, USA}
\author{Lewis~R.~B.~Picard}
\gradstudent
\author{Jessie~T.~Zhang}
\gradstudent
\author{William~B.~Cairncross}
\harvardccb
\harvardphysics
\cua
\author{Jeremy~M.~Hutson}
\affiliation{Joint Quantum Centre Durham-Newcastle, Department of Chemistry, Durham University, Durham, DH1 3LE, United Kingdom}
\author{Rosario Gonzalez-Ferez}
\affiliation{Instituto Carlos I de F\'{\i}sica Te\'orica y Computacional, and Departamento de F\'{\i}sica At\'omica, Molecular y Nuclear,  Universidad de Granada, 18071 Granada, Spain}
\itamp
\author{Till Rosenband}
\affiliation{Agendile LLC, Cambridge, Massachusetts 02139, USA}
\author{Kang-Kuen~Ni}
\email{ni@chemistry.harvard.edu}
\harvardccb
\harvardphysics
\cua

@@

In our system, the separation in energy between the initial and target state is much smaller than the single photon detuning $\Delta$ from the intermediate state; see Fig.~1a in the main text.
Thus, there is significant cross coupling for the scattering rate and light shift, where each state is coupled to the excited state by the power in both beams.
Under these conditions, the scattering rate and light shift are proportional to the total power $P_{\text{tot}} = P_{\rm m} + P_{\rm a}$, where $P_{\rm m} (P_{\rm a})$ is the power in the beam that addresses the molecular (atomic) state.
The Raman Rabi frequency is proportional to $\sqrt{P_{\rm m}P_{\rm a}} $.
With a fixed total power and thus fixed scattering rate, the Raman Rabi frequency is maximized when $ P_{\rm m} = P_{\rm a} = P_{\text{tot}}/2$.

@@

With multiple intermediate states, the total Raman Rabi frequency, scattering rate, and light shift become a sum over all these possible states. In our model, we consider all vibrational states of $ \mathrm{c^3\Sigma^+(\Omega = 1)} $ and states in the atomic continuum as intermediate states. The Rabi frequency between two states is proportional to the electronic dipole moment, arising from the electronic wavefunctions, and the overlap of the vibrational wavefunctions. Under the assumption that the electronic dipole moment $ d $ is the same for all transitions between the ground state and the excited electronic state, the total Raman Rabi frequency becomes

@@

  \Omega_{\rm R} = \frac{d^2E^2}{2\hbar^2} \displaystyle\sum_{v'} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}},

@@

where $\ket{a}$ ($\ket{m}$) are the relative coordinate wavefunctions for the trapped atomic (target molecular) state, and $\ket{v'}$ is the relative-coordinate wavefunction for the excited state which includes both bound molecular states and the atomic continuum. $\Delta_{v'}$ is a single photon detuning with respect to the $\ket{v'}$ state, $ E $ is the electric field for the power in the single beam given a waist size. We can divide this sum into a sum over all the bound molecular states, and a sum over atomic continuum states,

@@

  \Omega_{\rm R} = \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}} +  \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' > v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}},

@@

where $ v'_{\text{max}}$ corresponds to the excited molecular state, which is most weakly bound. The first term can be calculated numerically given an excited state potential. To calculate the second term, we first make an assumption that we are far enough detuned from the contributing continuum states that they are all at approximately the same detuning, $ \Delta_{\text{thresh}}$. This allows us to pull it out of the sum, and we obtain

@@

  \Omega_{\rm R} \approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}} +  \frac{d^2E^2}{2\hbar^2\Delta_{\text{thresh}}}\displaystyle\sum_{v' > v'_{\text{max}}} \braket{a|v'}\braket{v'|m} . \label{rabifreq}

@@

From requiring completeness and orthogonality,

@@

  \displaystyle\sum_{v'} \braket{a|v'}\braket{v'|m} = \braket{a|m} = 0,

@@

  \displaystyle\sum_{v'> v'_{\text{max}}} \braket{a|v'}\braket{v'|m} = -\displaystyle\sum_{v' \leq v'_{\text{max}}} \braket{a|v'}\braket{v'|m},

@@

  \Omega_{\rm R} \approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{a|v'}\braket{v'|m}}{\Delta_{v'}} -  \frac{d^2E^2}{2\hbar^2\Delta_{\text{thresh}}}\displaystyle\sum_{v' \leq v'_{\text{max}}} \braket{a|v'}\braket{v'|m}.

@@

We can account for the atomic continuum in the scattering rate in a similar way

@@

            \Gamma_{\rm s} &\approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}}\Gamma_{v'} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}^2} + \frac{d^2E^2\Gamma}{2\hbar^2\Delta_{\text{thresh}}^2}\displaystyle\sum_{v' > v'_{\text{max}}}\braket{m|v'}\braket{v'|m},

@@

  \displaystyle\sum_{v' > v'_{\text{max}}}\braket{m|v'}\braket{v'|m} = 1 - \displaystyle\sum_{v' \leq v'_{\text{max}}}\braket{m|v'}\braket{v'|m}.

@@

  \Gamma_{\rm s} \approx \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}}\Gamma_{v'} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}^2} +  \frac{d^2E^2\Gamma}{2\hbar^2\Delta_{\text{thresh}}^2}\left(1 - \displaystyle\sum_{v' \leq v'_{\text{max}}}\braket{m|v'}\braket{v'|m}\right).

@@

A very similar result holds for the light shift $\delta$ and is written here for completeness

\begin{equation}
  \delta = \frac{d^2E^2}{2\hbar^2}\displaystyle\sum_{v' \leq v'_{\text{max}}} \frac{\braket{m|v'}\braket{v'|m}}{\Delta_{v'}} +  \frac{d^2E^2}{2\hbar^2\Delta_{\text{thresh}}}\left(1 - \displaystyle\sum_{v' \leq v'_{\text{max}}}\braket{m|v'}\braket{v'|m}\right).
\end{equation}

\section{Computational methods for bound and scattering states}

@@

\section{Calculating Matrix Elements with Coupled Channel Ground State}
To calculate Rabi frequencies for transitions to the eight excited molecular potentials, we
obtain their wavefunctions from a single-channel calculation \cite{Fattal1996} using the potential energy curves described in section \ref{sm:excited_states}. We
then calculate overlap integrals between these functions and either the triplet ($S=1$) or singlet ($S=0$) components of the
coupled-channel wavefunctions for either the target molecular state or the confined atomic state.
These calculations assume that the intermediate states have pure singlet or triplet character and neglect its
nuclear spin character, which is unknown. These calculations include theoretical electronic transition dipole moments~\footnote{Private communication with R. Moszynski}.

@@

In this calculation, there is a matrix element, $\Omega_{\mathrm{a}ij}, \Omega_{\mathrm{m}ij}$ between each channel $ i $ of the coupled channel ground state and an excited state $ j $. The total Raman Rabi frequency $ \Omega_{\rm R}$, light shift $ \delta$ and off-resonant scattering $ \Gamma_{\rm s} $ contribution from this single excited state is calculated as

\begin{align}
  \Omega_{\mathrm{R}j} &= \displaystyle\sum_{i} \frac{\Omega_{\mathrm{a}ij} \Omega_{\mathrm{m}ij}}{2\Delta}, \\
  \delta_j &= \displaystyle\sum_{i} \frac{\Omega_{\mathrm{a}ij}^2 - \Omega_{\mathrm{m}ij}^2}{2\Delta}, \\
  \Gamma_{\mathrm{s} j} &= \displaystyle\sum_{i} \Gamma_{\rm e} \frac{\Omega_{\mathrm{a}ij}^2 + \Omega_{\mathrm{m}ij}^2}{2\Delta^2},
\end{align}
where $ \Delta $ is the single photon detuning and $ \Gamma_{\rm e} $ is the excited state linewidth. The experiment, through measuring the detuning dependence of the Raman Rabi frequency and the resonance frequency, extracts a single $ \Omega_{\rm m} $ and $ \Omega_{\rm a} $. In order to compare to the theory, we introduce effective $ \Omega_{\rm a}', \Omega_{\rm m}', \Gamma_{\rm e}'$, so that $ \Omega_{\rm R} = \Omega_{\rm a}'\Omega_{\rm m}'/(2\Delta) $, $\delta = (\Omega_{\rm a}'^2 - \Omega_{\rm m}'^2)/(2\Delta) $, and $\Gamma_{\rm s} = \Gamma_{\rm e}'(\Omega_{\rm a}'^2 + \Omega_{\rm m}'^2)/(2\Delta^2) $. The effective Rabi frequencies for the $v' = 0$ state of $ c^3\Sigma^+_1$ and their ratio are reported in Table~\ref{tab:sm} at $3.75~\mathrm{mW}$ of tweezer power.
